---
title: "OBDS Gene Set Enrichment and Pathway Analysis Practical"
author: "Charlie George"
date: "15/10/2021"
output: html_document
---

This is the pathway enrichment analysis practical of the OBDS course

- The first part of this pracical is using gprofiler to perform ORA on output dataset of differentially expressed genes from DESeq2

- The major aim of this exercise is to get you used to reading online tool documentation and playing with different paramters.

# load packages 
```{r}
# load librarys

library(tidyverse)
library(gprofiler2)
```


# Read in our tables from this morning in a new Rmd file
```{r}
# read in results table we output from DESeq2 this morning using tidyverse read_csv()

all_genes_df <- read_csv("~/0_git_repos/OBDS_Syllabus/courses/r/programme/differential_expression/CD8_vs_CD4_DE_results_lfcshrinkage.csv")

# check its dimensions - and look at the top of the file -> what does this tell us about our table? 
# dim = 22,535 -> this contains all genes -> need to filter to get those tested 
dim(all_genes_df)   
head(all_genes_df)
```

We want to perform a classic Over representation analsis (ORA) using gprofiler - to do this we first need to create our gene subsets 

```{r}

# Subset 1 - All genes tested in differential expression (this is our background/control gene list) - i.e. remove padj == NA 
all_genes_tested_df <- all_genes_df %>% 
                          dplyr::filter(padj !="NA")
  
  
# Subset 2 - All sig upregulated genes (padj < 0.05. & log2FoldChange > 1)
upreg_genes_df <- all_genes_tested_df %>% 
                    dplyr::filter(padj < 0.05 & log2FoldChange > 1)

# Subset 3 - All sig downregulated genes (padj < 0.05. & log2FoldChange < 1)
downreg_genes_df <- all_genes_tested_df %>% 
                    dplyr::filter(padj < 0.05 & log2FoldChange < -1)


# Check the dimensions of each of your dataframes using dim()
# all_genes_df_filtered should be 17729
# upreg_genes_df should be 915
# downreg_genes_df should be 1204
dim(all_genes_tested_df)
dim(upreg_genes_df)
dim(downreg_genes_df)

```


```{r}
# from each of these dataframes get a vector of the ensembl ids 
all_gene_ids <- all_genes_tested_df %>% pull(ensembl_gene_ids)
    
upreg_gene_ids <- upreg_genes_df %>% pull(ensembl_gene_ids)
    
downreg_gene_ids <- downreg_genes_df %>% pull(ensembl_gene_ids)
```

#Perform ORA using g:profiler 

1. Look at documentation - https://cran.r-project.org/web/packages/gprofiler2/vignettes/gprofiler2.html
```{r}
# use the gost() function - what paramters should we include/change 
# we want to test the ORA of upreg_gene_ids vs our all_gene_ids

# get help for function
?gost() 

# write query 
gost_results_obj <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", 
                ordered_query = FALSE,    
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = FALSE, # change to True as have alot of results otherwise - and we want to be stringent and look at well annotated genesets
                measure_underrepresentation = FALSE, 
                evcodes = FALSE, 
                user_threshold = 0.05, 
                correction_method = "g_SCS",  # this is thier own multiple testing coorrection - actually more strignent then bonferroni or BH 
                domain_scope = "annotated", # delete - will be set if have custom_bg automatically
                custom_bg = all_gene_ids, # THIS IS THE MOST IMPORTANT = this is our background gene set: make sure this is all gene ids that were tested in our differential expression
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = FALSE)

# what are our outputs  
head(gost_results_obj)


```

```{r}
# plot our results using in built function in gprofiler -> which one is it? 
gostplot(gost_results_obj, capped = TRUE, interactive = TRUE)
```

```{r}
# what does as_short_link = FALSE do in gost() function? 
gost_results_obj_link <- gost(query = upreg_gene_ids, 
                organism = "mmusculus", 
                ordered_query = FALSE,    
                multi_query = FALSE, 
                significant = TRUE, 
                exclude_iea = TRUE, # change to True as have alot of results otherwise
                measure_underrepresentation = FALSE, 
                evcodes = FALSE, 
                user_threshold = 0.05, 
                correction_method = "g_SCS",  # this is thier own multiple testing coorrection - actually more strignent then bonferroni or BH 
                custom_bg = all_gene_ids, # make sure this is all gene ids 
                numeric_ns = "", 
                sources = NULL, 
                as_short_link = TRUE)

gost_results_obj_link
```

Answer -> it gives you a url where you can quickly and easily visualize your data and send the link to others to see it too.  
  - point out `detailed results tab` 

```{r}
# Filter results table for genesets with between 5-300 genes 

gost_results_df = gost_results_obj$result
gost_results_df_5to300_genes = gost_results_df  %>% 
                                            filter(term_size >=5) %>% 
                                            filter(term_size <=300)

gost_results_df_5to300_genes

```


2. Try querying multiple gene lists at the same time (up and down regulated genes) using Multiple queries - see the documentation 

```{r}
multi_gostquery_results_obj <- gost(query = list("upreg_genes" = upreg_gene_ids,
                                                 "downreg_genes"= downreg_gene_ids),
                                    organism = "mmusculus", 
                                    exclude_iea = TRUE, # change to True as have a lot of results otherwise
                                    correction_method = "g_SCS",  # this is their own multiple testing correction - actually more stringent then bonferroni or BH 
                                    custom_bg = all_gene_ids, # make sure this is all gene ids 
                                    multi_query = TRUE)
```

Can you plot these querys together?

```{r}
gostplot(multi_gostquery_results_obj, capped = TRUE, interactive = TRUE)

```

Output the get_version_info() to save the versions of the databases that were used for the analysis!! 
```{r}
get_version_info('mmusculus')

```

Output the sessionInfo() so know what packages we used
```{r}
sessionInfo()

```

Now knit your report to get a lovely html file that you can share -> this is good practise as it creates a permanent record of your plots/analysis output and also makes sure your script runs without error from start to finish! 